Alone Together: Compositional Reasoning and Inference for Weak Isolation

### 2017-10-19 inference.tex
Notational changes
Also added proof sketch of theorems in Section 4 and 5

The correspondence between the inference rules and the RG judgment of
\S~\ref{sec:reasoning} is stated thus\footnote{The proof can be found in
the supplementary material.}:
\begin{theorem}
\label{thm:inference-sound}
  \emph{Forall} i,$R$,$I$,$c$,$\F$, if $\stable(\R,I)$ and $c \elabsto \F$,
  then:\\\vspace*{-0.2cm}
  \begin{smathpar}
  \begin{array}{c}
  \R \vdash \hoare{\lambda(\stl,\stg).~\stl=\emptyset \conj
  I(\stg)}{c}{\lambda(\stl,\stg).\stl = \F(\emptyset,\stg)}
  \end{array}
  \end{smathpar}
\end{theorem}
\noindent where the set expression $\F(\emptyset,\stg)$ has a natural
interpretation in the theory of sets.

\begin{proof}
  We only provide a brief proof sketch here. We prove a slightly
  stronger statement :

  \emph{Forall} i,$R$,$I$,$c$,$\F$,$s$ if $\stable(\R,I)$ and $c
  \elabsto \F$, then:\\

@@

  We use structural induction on $c$ to prove the statement. Assume
  that $P \Leftrightarrow \lambda(\stl,\stg).~\stl=s \conj I(\stg)$
  and $Q \Leftrightarrow \lambda(\stl,\stg).\stl = \F(s,\stg)$. For
  each type of $c$, our strategy is to show the premises of the
  corresponding RG-rule. The base cases correspond to the INSERT,
  UPDATE and DELETE statements. In all these cases, the form of $F$
  directly corresponds to the condition linking the pre-condition $P$
  and post-condition $Q$ in the premises of the RG-rules. If
  $\stable(\R, F)$, then $\stabilize{F} = F$ and hence the proof
  directly follows. Otherwise, $\stabilize{F} = \lambda (\stl,
  \stg).\existsl(\stg', I(\stg'), F(\stl, \stg'))$. Since the
  pre-condition $P(\stl, \stg)$ asserts $I(\stg)$, any state which
  obeys $P$ provides the existential $\stg'$ in the above set
  expression, and hence $F$ applied on that state matches the
  condition linking $P$ and $Q$. The stability of $P$ with respect to
  $\R$ follows directly from the stability of $I$.

  The proofs for SELECT, SEQ and IF-THE-ELSE directly use the inductive
  hypothesis to infer the RG-judgements present in the premises of
  their corresponding RG-rules. The interesting case is the FOREACH
  statement, for which we use the loop invariant $\psi(\stl, \stg)
  \Leftrightarrow \stl = s \cup y \bind (\lambda z.~\F(s, \stg))$,
  (where assuming that $c$ is the body of the loop, $c \elabsto F$) to
  prove all the premises of \rulelabel{RG-Foreach}. Using the same
  notation as the rule $\rulelabel{RG-Foreach}$, $y$ refers to the
  records already processed in previous iterations of the loop, while
  $z$ refers to the record being processed in the current iteration.
  At the beginning of the loop $[\phi/y]\psi(\stl, \stg)$ just reduces
  to $\stl = s$ which is implied by the pre-condition $P$. From the
  inductive hypothesis, we can infer that each iteration corresponds
  to the application of $F$. Since all iterations are assumed to be
  independent of each other, and $z$ is bound to a record in $x$ for
  each iteration, we can infer that at the end of every iteration the
  loop invariant $[y \cup \{z\}/y]\psi$ will be satisfied. 
\end{proof}

@@

Theorem~\ref{thm:inference-sound} lets us replace the local judgment
of the \rulelabel{RG-Txn} rule (Fig.~\ref{fig:rg-rules}) by a state
transformer inference judgment. The soundness of a transaction's
guarantee can now be established w.r.t the effect $\F$ of the body.
The \rulelabel{RG-Txn} rule so updated is shown below:
\begin{smathpar}
\begin{array}{c}
\RULE
{
  \stable(R,\I)\spc
  \stable(R,I)\spc
  \R_e(\stl,\stg,\stg') \Leftrightarrow \exists \stg_1.  R(\stg, \stg') \wedge \I_e(\stl, \stg_1, \stg') \spc  c \Longrightarrow_{\langle i,\R_e,I \rangle}\F\\
  \R_c(\stl,\stg,\stg') \Leftrightarrow \exists \stg_1.  R(\stg, \stg') \wedge \I_c(\stl, \stg_1, \stg') \spc \stable(\R_c,\F)\\
  \forall \stl,\stg.~ \stl \in F(\emptyset,\stg) \Rightarrow 
    G(\stg, \stl\gg\stg)\spc
  \forall \stg,\stg'.~I(\stg) \wedge G(\stg,\stg') \Rightarrow I(\stg')\\
}
{
  \rg{I,R}{\ctxn{i}{\I}{c}}{G,I}
}
\end{array}
\end{smathpar}
Automating the application of the \rulelabel{RG-Txn} rule for a
transaction requires automating the multiple implication checks in
the premise. While $R$, $\R$, $\R_c$, $\I$ and $I$ are formulas in
first-order logic (FOL) with a relatively simple structure, $\F$
is an expression in the set language $\SL$
(Fig.~\ref{fig:logic-syntax}) with a possibly complex structure.
Fortunately, however, there exists a semantics-preserving translation
from $\SL$ to a restricted subset of first-order logic (FOL) that
lends itself to automatic reasoning.

>>>>

The correspondence between the inference rules and the RG judgment of
\S~\ref{sec:reasoning} is stated thus:
\begin{theorem}
\label{thm:inference-sound}
  \emph{Forall} i,$R$,$I$,$c$,$\F$, if $\stable(\R,I)$ and $c \elabsto \F$,
  then:\\\vspace*{-0.2cm}
  \begin{smathpar}
  \begin{array}{c}
  \R \vdash \hoare{\lambda(\stl,\stg).~\stl=\emptyset \conj
  I(\stg)}{c}{\lambda(\stl,\stg).\stl = \F(\emptyset,\stg)}
  \end{array}
  \end{smathpar}
\end{theorem}
\noindent where the set expression $\F(\emptyset,\stg)$ has a natural
interpretation in the theory of sets.

\begin{proof}
  We only provide a brief proof sketch here\footnote{The complete proof can be found in
the supplementary material.}. We prove a slightly
  stronger statement :

  \emph{Forall} i,$R$,$I$,$c$,$\F$,$s$ if $\stable(\R,I)$ and $c
  \elabsto \F$, then:

@@

  We use structural induction on $c$. Assume
  that $P \Leftrightarrow \lambda(\stl,\stg).~\stl=s \conj I(\stg)$
  and $Q \Leftrightarrow \lambda(\stl,\stg).\stl = \F(s,\stg)$. For
  each type of $c$, our strategy is to show the premises of the
  corresponding RG-rule. The base cases correspond to the INSERT,
  UPDATE and DELETE statements. In all these cases, the form of $F$
  directly corresponds to the condition linking the pre-condition $P$
  and post-condition $Q$ in the premises of the RG-rules. If
  $\stable(\R, F)$, then $\stabilize{F} = F$ and hence the proof
  directly follows. Otherwise, $\stabilize{F} = \lambda (\stl,
  \stg).\existsl(\stg', I(\stg'), F(\stl, \stg'))$. Since the
  pre-condition $P(\stl, \stg)$ asserts $I(\stg)$, any state which
  obeys $P$ provides the existential $\stg'$ in the above set
  expression, and hence $F$ applied on that state matches the
  condition linking $P$ and $Q$. The stability of $P$ with respect to
  $\R$ follows directly from the stability of $I$.

  The proofs for SELECT, SEQ and IF-THEN-ELSE directly use the inductive
  hypothesis to infer the RG-judgements present in the premises of
  their corresponding RG-rules. The interesting case is the FOREACH
  statement, for which we use the loop invariant $\psi(\stl, \stg)
  \Leftrightarrow \stl = s \cup (y \bind (\lambda z.~\F(s, \stg)))$,
  (where assuming that $c$ is the body of the loop, $c \elabsto F$) to
  prove all the premises of \rulelabel{RG-Foreach}. Using the same
  notation as the rule $\rulelabel{RG-Foreach}$, $y$ refers to the
  records already processed in previous iterations of the loop, while
  $z$ refers to the record being processed in the current iteration.
  At the beginning of the loop $[\phi/y]\psi(\stl, \stg)$ just reduces
  to $\stl = s$ which is implied by the pre-condition $P$. From the
  inductive hypothesis, we can infer that each iteration corresponds
  to the application of $F$. Since all iterations are assumed to be
  independent of each other, and $z$ is bound to a record in $x$ for
  each iteration, we can infer that at the end of every iteration the
  loop invariant $[y \cup \{z\}/y]\psi$ will be satisfied. 
\end{proof}

@@

Theorem~\ref{thm:inference-sound} lets us replace the local judgment
of the \rulelabel{RG-Txn} rule (Fig.~\ref{fig:rg-rules}) by a state
transformer inference judgment. The soundness of a transaction's
guarantee can now be established w.r.t the effect $\F$ of the body.
The \rulelabel{RG-Txn} rule so updated is shown below:
\begin{smathpar}
\begin{array}{c}
\RULE
{
  \stable(R,\I)\spc
  \stable(R,I)\spc
  P(\stl,\stg) \Leftrightarrow \stl=\emptyset \wedge I(\stg)\\
  \R_e = R \backslash \I_e \spc \R_c = R \backslash \I_c \spc 
   \R_e \vdash \rg{P}{c}{Q} \spc \stable(\R_c,Q) \\ 
  \forall \stl,\stg.~ Q(\stl,\stg) \Rightarrow 
    G(\stg, \stl \rhd \stg)\spc
  \forall \stg,\stg'.~I(\stg) \wedge G(\stg,\stg') \Rightarrow I(\stg')\\
}
{
  \rg{I,R}{\ctxn{i}{\I}{c}}{G,I}
}
\end{array}
\end{smathpar}
Automating the application of the \rulelabel{RG-Txn} rule for a
transaction requires automating the multiple implication checks in
the premise. While $R$, $\R$, $\R_c$, $\I$ and $I$ are formulas in
first-order logic (FOL) with a relatively simple structure, $\F$
is an expression in the set language $\SL$
(Fig.~\ref{fig:logic-syntax}) with a possibly complex structure.
Fortunately, however, there exists a semantics-preserving translation
from $\SL$ to a restricted subset of first-order logic (FOL) that
lends itself to automatic reasoning.

### 2017-10-19 reasoning.tex
Notational changes
Also added proof sketch of theorems in Section 4 and 5

A characteristic of RG reasoning is that stability of an assertion is
always proven w.r.t to $R$, and not $R^{*}$, although interference may
include multiple environment steps, and $R$ only captures a single
step. This is nonetheless sound due to inductive reasoning: if $\phi$
is preserved by every step of $R$, then $\phi$ is preserved by
$R^{*}$, and vice-versa.  However, such reasoning does not extend
naturally to isolation-constrained interference because $R^{*}$ modulo
$\I$ is not same as $\R^{*}$; the former is a transitive relation
constrained by $\I$, whereas the latter is the transitive closure of
an This means, unfortunately, that we cannot directly replace $R^{*}$
by $R$ in the above condition because of our need to consider the weak
isolation specification $\I$ that selectively exposes interferences.

@@

\noindent We can now assert the soundness of the RG judgments in
Fig.~\ref{fig:rg-rules} as follows\footnote{Its proof is provided in
  the supplementary material.}:

@@

\noindent Thus, if $\rg{I,R}{c}{G,I}$ is a valid RG judgment, then (a)
every interleaved multi-step reduction of $c$ preserves the database
integrity constraint (consistency condition) $I$, and (b) the effect
that every transaction in $c$ has on the database state is captured by
$G$.

>>>>

A characteristic of RG reasoning is that stability of an assertion is
always proven w.r.t to $R$, and not $R^{*}$, although interference may
include multiple environment steps, and $R$ only captures a single
step. This is nonetheless sound due to inductive reasoning: if $\phi$
is preserved by every step of $R$, then $\phi$ is preserved by
$R^{*}$, and vice-versa.  However, such reasoning does not extend
naturally to isolation-constrained interference because $R^{*}$ modulo
$\I$ is not same as $\R^{*}$; the former is a transitive relation
constrained by $\I$, whereas the latter is the transitive closure of
a constrained relation. This means, unfortunately, that we cannot directly replace $R^{*}$
by $R$ in the above condition because of our need to consider the weak
isolation specification $\I$ that selectively exposes interferences.

@@

\noindent Thus, if $\rg{I,R}{c}{G,I}$ is a valid RG judgment, then (a)
every interleaved multi-step reduction of $c$ preserves the database
integrity constraint (consistency condition) $I$, and (b) the effect
that every transaction in $c$ has on the database state is captured by
$G$.
\noindent We can now assert the soundness of the RG judgments in
Fig.~\ref{fig:rg-rules} as follows:

@@

\begin{proof}
Here, we provide only a brief proof sketch\footnote{The complete proof is provided in the supplementary material.}. The most important rule is the top-level rule \rulelabel{RG-Txn}, which proves that a transaction $c$ which begins its execution in global database state satisfying $I$ and suffering interference $R$ while executing under isolation specification $\I$ finishes its execution in a database state also satisfying $I$, and also guarantees that its commit step satisfies $G$. The rule uses the transaction-local RG judgement $\R_e \vdash \rg{P}{c}{Q}$. By \rulelabel{E-Txn-Start}, the local and global database states at the start of a transaction satisfy $P$, and the only challenge is that environment steps in an execution covered by $\R_e \vdash \rg{P}{c}{Q}$ are in $\R_e$, while the top-level judgement requires environment steps in $R$. We show that it is enough to consider only those enviroment steps in $\R_e$. First, we use an inductive argument and stability of $\I_e$ ($\stable(R, \I_e)$) to show that any execution in which the transaction completes all its steps must always preserve the isolation specification $\I_e$ after every environment step. Intuitively, this is because once $\I_e$ gets broken after some environment step, it will continue to remain broken and the transaction would not be able to proceed (according to \rulelabel{E-Txn}). Since $\R_e$ contains exactly those environment steps which preserve $\I_e$, the local-level RG judgement can be soundly used, which guarantees that after the transaction finishes its execution, its local state $\stl$ and global state $\stg$ will satisfy the assertion $Q$. Environment steps between the last step of the transaction and its commit step can modify the global state, and hence we also require $Q$ to be stable againt $R$. Again, we use an inductive argument, the stability of $\I_c$, and the fact that the transaction must execute its commit step to show that all environment steps must preserve $\I_c$, and hence it is enough to require $\stable(\R_c, Q)$. $Q$ guarantees that the commit step is in $G$, and $G$ in turn guarantees that after execution, the global database state will obey the invariant $I$.

The local RG-judgement rules are mostly straightforward and directly follow from the operational semantics. At this point, the transaction can only change its local database state by adding new records to it, and hence all the rules have the form : $P(\stl, \stg)$ and $\stl'$ is the new state should satisfy $Q(\stl', \stg)$. An important thing to note is that the rules for Update, Select, Delete and Insert only require the pre-condition $P$ to be stable against interference. This is because the local RG-judgement rules only make assertions about the database state just after the transaction takes its last step. For the aforementioned commands, the last step would be the only step taken by the transaction, and hence $Q$ would not be affected by any environment steps, but instead directly reflect the changes made by the command (the environment steps between the last execution step of the transaction and the commit step would be controlled by $\I_c$ and would be taken care of by $\stable(\R_c, Q)$ at the top level).

The rule $\rulelabel{RG-Foreach}$ makes use of an inductive loop invariant $\psi$ which assumes that the variable $y$ is bound to records which have been processed in previous iterations and $z$ is bound to the record processed in the current iteration. Since the operational semantics ensure that every iteration processes a unique record, the proof is straightforward.  
\end{proof}
